---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# tabfm

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

The goal of tabfm is to make [TabFM](https://github.com/google-research/tabfm),
Google Research's zero-shot foundation model for tabular data, easy to use
from R. You give it a data frame and a response; it gives you classification
or regression predictions with **no feature engineering, no hyperparameter
tuning, and no retraining** — the model predicts unseen tables in-context, in
a single forward pass. See the
[announcement blog post](https://research.google/blog/introducing-tabfm-a-zero-shot-foundation-model-for-tabular-data/)
for details on how it works.

tabfm wraps the official Python package with
[reticulate](https://rstudio.github.io/reticulate/), and exposes the familiar
R modelling interface: a fitting function plus `predict()`.

## Installation

You can install the development version of tabfm like so:

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("mattyoreilly/tabfm")
```

Then, one time only, install the underlying Python package into a dedicated
virtualenv:

```{r, eval = FALSE}
library(tabfm)
install_tabfm()            # JAX backend, CPU
# install_tabfm("pytorch") # PyTorch backend, CPU/GPU
```

## Example

Fitting is instant — TabFM learns your table in-context rather than training
on it — and the loaded model stays in memory, so repeated predictions are
cheap. Character and factor columns are handled natively; you never need to
dummy-code.

```{r example}
library(tabfm)

fit <- tabfm(iris[-5], iris$Species)
fit

predict(fit, head(iris[-5]))

predict(fit, head(iris[-5]), type = "prob")
```

Regression works the same way — pass a numeric response and
`task = "regression"`:

```{r regression}
fit <- tabfm(mtcars[-1], mtcars$mpg, task = "regression")
predict(fit, head(mtcars[-1]))
```

## Making it fast

The pretrained weights are cached after the first download, but two optional
lines in your `~/.Renviron` (no Hugging Face account needed) make cold starts
noticeably faster:

```
HF_HUB_OFFLINE=1
JAX_COMPILATION_CACHE_DIR=~/.cache/jax
```

* `HF_HUB_OFFLINE=1` loads weights straight from the local cache instead of
  revalidating them against the Hugging Face Hub on each session's first
  fit — unauthenticated requests get the slowest rate-limit tier, so this
  skips the wait entirely. Unset it whenever you need to download weights
  you don't have yet.
* `JAX_COMPILATION_CACHE_DIR` lets new sessions reuse the compiled XLA
  artifacts from previous ones instead of recompiling the model graph.

Within a session, the loaded model is cached automatically: only the first
`tabfm()` call per task pays the load cost.

## Limitations

* Predictions require a working Python installation; `install_tabfm()` sets
  everything up via reticulate.
* The first `tabfm()` call downloads the pretrained weights from the
  Hugging Face Hub (cached afterwards in `~/.cache/huggingface/`). You may
  see a warning about unauthenticated requests; it's harmless, but you can
  silence it and get faster downloads by putting a free Hugging Face
  [access token](https://huggingface.co/settings/tokens) in your
  `~/.Renviron`: `HF_TOKEN=hf_...`
* TabFM is zero-shot: there is nothing to tune, but also no way to trade
  compute for accuracy on a specific dataset. If you need that, reach for
  [tidymodels](https://www.tidymodels.org).
